Sequential rank agreement methods for comparison of lists

### 2015-08-09 latex/paper.Rnw
More comment fixing

Each spectrum consists of $49642$ samples over a range of mass-to-charge ratios
between $800$ to $20000$ Dalton which we downsample on an equidistant grid of
5000 points by linear interpolation. We then preprocess the downsampled
spectra individually by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. We then standardize
the 5000 predictors to have column-wise zero mean and unit variance in
each data set.

We use the two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of four different statistical
classification methods in how they rank the predictors according to
their importance for distinguishing benign and malignant tumors. In
the second scenario we assess the agreement among rankings of
individual predicted risks of having a malignant tumor. The first
scenario is relevant in the context of biomarker discovery and the
latter is important, e.g., to rank patients according to immediacy of
treatment.

@@

All four methods depend on a tuning parameter. We select these
separately for each sub-sample by a 20 times repeated 5-fold
cross-validation procedure. The tuning parameter for Lasso and Ridge regression
is the degree of penalization, for PLS-DA it is the number of components
(the dimensionality of the subspace), and for Random Forest it is the number
 of predictors randomly sampled at each split.
The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce censored lists depending on
the value of the penalty parameter.

\subsection{Agreement of predictor rankings}
For each of the four methods, each of the 1000 models trained in on of
the 1000 sub-samples of the MALOVA data produces a ranking of the 5000
predictors according to their importance for discriminating between
the tumor types. For the Random Forest classifier the predictors are ranked according
to the Gini index, while for the logistic Lasso and Ridge regression
models we order the predictors by absolute magnitude of the estimated
regression coefficients. For the PLS-DA model the importance of the
predictors is based on a weighted sum of the absolute coefficients
where the weights are proportional to the reduction in the sums of
squares across the components.

The right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
600. At deeper list depths all agreement curves become approximately constant.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
\caption{Left panel: Sequential rank agreement of 1000 rankings of the
  predicted risks of malignant tumor. For each method the different
  rankings were obtained by first training models in 1000 random
  sub-samples of the MALOVA data and then predicting the risk of
  malignant tumor in the 113 MALOVA patients. Right panel: Sequential
  rank agreement of 1000 rankings of the 5000 predictors. The rankings
  were obtained from the same 1000 trained models.}
 \label{fig:app1}
\end{figure}

@@

Most curves start low and then increase indicating higher agreement
among high risk patients. This is expected if we rank the individuals
according to highest risk of disease. However, it is also expected
that individuals with very low risk also show high agreement. In this
case we order the patients according to (high) risk prediction but we
could essentially also have reversed the order to identify the
patients that have low risk prediction.  An exception is the risk
prediction agreement for the Lasso tuned with the AUC criterion which
shows very low agreement among the high values of the predicted
risks. The reason is that optimizing the penalty parameter with
respect to the AUC criterion tends to favor a very high penalty value
causing only a single predictor to be selected in each of the 1000
iterations. This results in a lack of generalizability to the DACOVA
data which gives rise to the higher disagreement in the predicted risk
probabilities. In the extreme case where the penalty becomes so high
that none of the predictors are selected by the Lasso, the sequential
rank agreement for the predicted probabilities becomes undefined since
all the ranks will be ties.

@@

In Figure \ref{fig:app1} it can further be seen that some of the
methods show better agreement with respect to the predicted
probabilities than for ranking the importance of the predictors and
vice versa. Ridge regression shows higher agreement across training
sets for the risk predictions than PLS-DA, and PLS-DA shows higher
agreement for predictor importance than Ridge regression. The Random
Forest method seems to perform equally well in both scenarios.
\textbf{Andreas: jeg forstaar ikke den sidste saetning. kan man sammenligne de to rank agreements (left vs right panel)?}
Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poor agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the Lasso penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable is selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.71$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
\textbf{and for this data set it appears that PLS-DA is the overall best
  method taking both agreement and discrimination ability into account.
  Andreas: vi boer ogsaa se paa Brier score (calibration) inden vi konkluderer at PLS-DA er den bedste
}

>>>>

Each spectrum consists of $49642$ samples over a range of mass-to-charge ratios
between $800$ to $20000$ Dalton which we downsample on an equidistant grid of
5000 points by linear interpolation. We then preprocess the downsampled
spectra individually by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. Finally, we standardize
the $5000$ predictors to have column-wise zero mean and unit variance in
each data set.

We use the two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of four different statistical
classification methods in how they rank the predictors according to
their importance for distinguishing benign and malignant tumors. In
the second scenario we assess the agreement among rankings of
individual predicted risks of having a malignant tumor. The first
scenario is relevant in the context of biomarker discovery and the
latter is important e.g., when ranking patients according to immediacy of
treatment.

@@

All four methods depend on a tuning parameter. The tuning parameter for 
Lasso and Ridge regression is the degree of penalization, and for PLS-DA 
it is the number of components (the dimensionality of the subspace). 
We estimate these separately for each sub-sample by a 20 times repeated 
5-fold cross-validation procedure. For the Random Forest we grow a fixed
number of $5000$ trees and let the tuning parameter be the number of 
predictors randomly sampled at each split. We estimate this by a binary
search with respect to minimizing the Out-of-Bag classification error estimate.

The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce censored lists depending on
the value of the penalty parameter.

\subsection{Agreement of predictor rankings}
For each of the four methods, each of the 1000 models trained on 
the 1000 sub-samples of the MALOVA data produces a ranking of the 5000
predictors according to their importance for discriminating between
the tumor types. For the Random Forest classifier the predictors are ranked according
to the Gini index, while for the logistic Lasso and Ridge regression
models we order by absolute magnitude of the estimated
regression coefficients. For the PLS-DA model the importance of the
predictors is based on a weighted sum of the absolute coefficients
where the weights are proportional to the reduction in the sums of
squares across the components.

The right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
600. At deeper list depths all agreement curves are approximately constant.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
\caption{Left panel: Sequential rank agreement of 1000 rankings of the
  predicted risks of malignant tumor. For each method the different
  rankings were obtained by first training models in 1000 random
  sub-samples of the MALOVA data and then predicting the risk of
  malignant tumor in the 113 DACOVA patients. Right panel: Sequential
  rank agreement of 1000 rankings of the 5000 predictors. The rankings
  were obtained from the same 1000 trained models.}
 \label{fig:app1}
\end{figure}

@@

Most curves start low and then increase indicating higher agreement
among high risk patients. This is expected if we rank the individuals
according to highest risk of disease. However, it is also expected
that individuals with very low risk also show high agreement. In this
case we order the patients according to (high) risk prediction but we
could essentially also have reversed the order to identify the
patients that have low risk prediction.  An exception is the risk
prediction agreement for the Lasso tuned with the AUC criterion which
shows very low agreement among the high values of the predicted
risks. The reason is that optimizing the penalty parameter with
respect to the AUC criterion tends to favor a very high penalty value
causing only a single predictor to be selected in each of the 1000
iterations. This results in a lack of generalizability to the DACOVA
data which gives rise to the higher disagreement in the predicted risks. 
In the extreme case where the penalty becomes so high
that none of the predictors are selected by the Lasso, the sequential
rank agreement for the predicted probabilities becomes undefined since
all the ranks will be ties.

@@

Comparing the left and right panels of Figure \ref{fig:app1} it can 
further be seen that some of the methods show better agreement with 
respect to the predicted probabilities than for ranking the importance 
of the predictors and vice versa. Ridge regression shows higher agreement across training
sets for the risk predictions than PLS-DA, and PLS-DA shows higher
agreement for predictor importance than Ridge regression.

Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poor agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the Lasso penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable is selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC and Brier scores for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.71$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
\textbf{Comment on Brier scores}
Both concerns are of course important in practice and for this data set 
it appears thatPLS-DA is the overall best method taking both agreement 
and classification accuracy into account.
